Viral Sequence Sorting Evaluation

### 2022-07-01 Scripts/Evaluations/EarlyIterations/identifying_viral_contigs_20220515.Rmd
Reorganizing scripts.

#vibrant
viruses$keep_score_vibrant[viruses$Quality=="high quality draft"] <- 1
viruses$keep_score_vibrant[viruses$Quality=="medium quality draft"] <- 0.52
viruses$keep_score_vibrant[viruses$Quality=="low quality draft"] <- 0.333
viruses$keep_score_vibrant[viruses$viral_genes==0 & viruses$host_genes>=1] <- 0
viruses$keep_score_vibrant[(viruses$viral_genes*3) <= viruses$host_genes] <- 0
viruses$keep_score_vibrant[viruses$contig_length>50000 & viruses$hallmark==0] <- 0

@@

#virsorter2
viruses$keep_score_virsorter2[viruses$viral>=50] <- 0.333
viruses$keep_score_virsorter2[viruses$hallmark>2] <- 1
viruses$keep_score_virsorter2[viruses$viral>=95] <- viruses$keep_score_virsorter2[viruses$viral>=95] + 1
viruses$keep_score_virsorter2[viruses$viral_genes==0 & viruses$host_genes>=1] <- 0
viruses$keep_score_virsorter2[(viruses$viral_genes*3) <= viruses$host_genes] <- 0
viruses$keep_score_virsorter2[viruses$contig_length>50000 & viruses$hallmark==0] <- 0

@@

>>>>

#vibrant
viruses$keep_score_vibrant[viruses$Quality=="high quality draft"] <- 1
viruses$keep_score_vibrant[viruses$Quality=="medium quality draft"] <- 0.52
viruses$keep_score_vibrant[viruses$Quality=="low quality draft"] <- 0.52
viruses$keep_score_vibrant[viruses$viral_genes==0 & viruses$host_genes>=1] <- 0
viruses$keep_score_vibrant[(viruses$viral_genes*3) <= viruses$host_genes] <- 0
viruses$keep_score_vibrant[viruses$contig_length>50000 & viruses$hallmark==0] <- 0

@@

#virsorter2
viruses$keep_score_virsorter2[viruses$viral>=50] <- 0.52
viruses$keep_score_virsorter2[viruses$hallmark>2] <- 1
viruses$keep_score_virsorter2[viruses$viral>=95] <- viruses$keep_score_virsorter2[viruses$viral>=95] + 1
viruses$keep_score_virsorter2[viruses$viral_genes==0 & viruses$host_genes>=1] <- 0
viruses$keep_score_virsorter2[(viruses$viral_genes*3) <= viruses$host_genes] <- 0
viruses$keep_score_virsorter2[viruses$contig_length>50000 & viruses$hallmark==0] <- 0

@@

# Visualizing which tools led to a contig being selected
(alluvial plot)
with help from: https://www.bioinfo-scrounger.com/archives/alluvial-and-Sankey/

```{r}
alluvial_data <- viruses %>% select(keep_score_vibrant:keep_score_checkV) %>%
  count(keep_score_vibrant, keep_score_virsorter2, keep_score_deepvirfinder, keep_score_checkV)
```

```{r}
alluvial_data$cum_sum <- alluvial_data %>% select(keep_score_vibrant:keep_score_checkV) %>% 
  rowSums()
```

```{r}
alluvial_data$viral <- "not viral"
alluvial_data$viral[alluvial_data$cum_sum>=1] <- "one method"
alluvial_data$viral[alluvial_data$cum_sum>=2] <- "two methods"
alluvial_data$viral[alluvial_data$cum_sum>=3] <- "three methods"
alluvial_data$viral[alluvial_data$cum_sum>=4] <- "all methods"

alluvial_data$viral <- factor(alluvial_data$viral, levels=c("not viral", "one method",
                                                            "two methods", "three methods",
                                                            "all methods"))
```

```{r}
alluvial_data <- alluvial_data %>% 
  mutate(across(keep_score_vibrant:keep_score_checkV, as.factor))
```

```{r}
head(alluvial_data)
```

```{r}
library(ggalluvial)
```

```{r}
ggplot(data = alluvial_data,
       aes(axis1 = as.character(keep_score_vibrant), 
           axis2 = as.character(keep_score_virsorter2),
           axis3 = as.character(keep_score_deepvirfinder),
           axis4 = as.character(keep_score_checkV),
           y = n)) +
  geom_alluvium(aes(fill = viral)) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum))) +
  scale_fill_viridis_d() + scale_x_discrete(limits = c("VIBRANT", "VirSorter2", "DeepVirFinder", "CheckV"),
                                  expand = c(0.05, 0.05, 0.05, 0.05)) +
  ylab("Number of Sequences") +
  theme_classic()
```

#########################################################
# Slightly modified rules for visualization purposes

#vibrant
viruses$keep_score_vibrant[viruses$Quality=="high quality draft"] <- 1
viruses$keep_score_vibrant[viruses$Quality=="medium quality draft"] <- 0.5
viruses$keep_score_vibrant[viruses$Quality=="low quality draft"] <- 0.5

#deepvirfinder
viruses$keep_score_deepvirfinder[viruses$bh_pvalue<=0.05] <- 1

#virsorter2
viruses$keep_score_virsorter2[viruses$viral>=50] <- 0.5
viruses$keep_score_virsorter2[viruses$hallmark>2] <- 1
viruses$keep_score_virsorter2[viruses$viral>=95] <- viruses$keep_score_virsorter2[viruses$viral>=95] + 1

#checkV
viruses$keep_score_checkV[(viruses$viral>=50 & viruses$host_genes==0) & viruses$viral_genes==0] <- viruses$keep_score_checkV[(viruses$viral>=50 & viruses$host_genes==0) & viruses$viral_genes==0] + 0.5
viruses$keep_score_checkV[viruses$percent_viral>=50] <- viruses$keep_score_checkV[viruses$percent_viral>=50] + 1
viruses$keep_score_checkV[viruses$viral_genes==0 & viruses$host_genes>=1] <- -5
viruses$keep_score_checkV[(viruses$viral_genes*3) <= viruses$host_genes] <- -5
viruses$keep_score_checkV[viruses$contig_length>50000 & viruses$hallmark==0] <- -5
```

# Visualizing which tools led to a contig being selected
(alluvial plot)
with help from: https://www.bioinfo-scrounger.com/archives/alluvial-and-Sankey/

```{r}
alluvial_data <- viruses %>% select(keep_score_vibrant:keep_score_checkV) %>%
  count(keep_score_vibrant, keep_score_virsorter2, keep_score_deepvirfinder, keep_score_checkV)
```

```{r}
alluvial_data$cum_sum <- alluvial_data %>% select(keep_score_vibrant:keep_score_checkV) %>% 
  rowSums()
```

```{r}
alluvial_data$viral <- "not viral"
alluvial_data$viral[alluvial_data$cum_sum>=1] <- "one method"
alluvial_data$viral[alluvial_data$cum_sum>=2] <- "two methods"
alluvial_data$viral[alluvial_data$cum_sum>=3] <- "three methods"
alluvial_data$viral[alluvial_data$cum_sum>=4] <- "all methods"

alluvial_data$viral <- factor(alluvial_data$viral, levels=c("not viral", "one method",
                                                            "two methods", "three methods",
                                                            "all methods"))
```

```{r}
alluvial_data <- alluvial_data %>% 
  mutate(across(keep_score_vibrant:keep_score_checkV, as.factor))
```

```{r}
head(alluvial_data)
```

```{r}
ggplot(data = alluvial_data,
       aes(axis1 = as.character(keep_score_vibrant), 
           axis2 = as.character(keep_score_virsorter2),
           axis3 = as.character(keep_score_deepvirfinder),
           axis4 = as.character(keep_score_checkV),
           y = n)) +
  geom_alluvium(aes(fill = viral)) +
  geom_stratum() +
  geom_text(stat = "stratum",
            aes(label = after_stat(stratum))) +
  scale_fill_viridis_d() + scale_x_discrete(limits = c("VIBRANT", "VirSorter2", "DeepVirFinder", "CheckV"),
                                  expand = c(0.05, 0.05, 0.05, 0.05)) +
  ylab("Number of Sequences") +
  theme_classic()
```

=> once have the testing set data, make this figure but split by type of sequence

# Other visualizations of output

### 2022-07-01 Scripts/Evaluations/EarlyIterations/viral_contig_sorting_testing_set_20220521_to_share.Rmd
Reorganizing scripts.

## Kaiju
(parsing when reading the file in "fails" for sequences that weren't classified)
```{r}
kj_c <- read_tsv("../VSTE/ToolOutput/metagenomic_testing_set_nrandeuk.kaiju.names.out", col_names = F)
colnames(kj_c) <- c("Classified", "Contig", "NCBI_taxon", "len", "ID_best", "IDs_all", "Seq", "Name")
kj_c <- separate(kj_c, col = Contig, into = c("seqtype", "contig"), sep="--")
kj_c <- separate(kj_c, col = Name, into = c("Kaiju_Viral","Kingdom"), sep=";")
kj_c$contig <- sub("\\.", "_", kj_c$contig)
```

@@

## VirSorter
```{r}
vs_c <- read_csv("../TestingSetOutput/VIRSorter_global-phage-signal.csv")
vs_c$`## Contig_id` <- sub("VIRSorter_", "", vs_c$`## Contig_id`)
vs_c <- separate(vs_c, col = `## Contig_id`, into = c("seqtype", "contig"), sep="--")
vs_c$contig <- sub("-circular", "", vs_c$contig)
vs_c <- vs_c[!duplicated(vs_c$contig),]
```

@@

```{r}
viruses <- full_join(x=vs2_c, y=dvf_c, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=checkV, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=vb_c, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=vs_c, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=kj_c, by = c("contig", "seqtype"))
```

@@

get rid of NAs for downstream processing
```{r}
viruses$Quality[is.na(viruses$Quality)] <- 0
viruses$score[is.na(viruses$score)] <- 0
viruses$bh_pvalue[is.na(viruses$bh_pvalue)] <- 0
viruses$viral[is.na(viruses$viral)] <- 0
viruses$hallmark[is.na(viruses$hallmark)] <- 0
viruses$percent_viral[is.na(viruses$percent_viral)] <- 0
viruses$percent_unknown[is.na(viruses$percent_unknown)] <- 0
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
viruses$Kingdom[is.na(viruses$Kingdom)] <- "unknown"
viruses$Category[is.na(viruses$Category)] <- 0
```

@@

```{r}
getting_viral_set_1 <- function(input_seqs, include_vibrant=FALSE, include_virsorter2=FALSE,
                              include_deepvirfinder=FALSE, include_checkV=FALSE,
                              include_kaiju=FALSE, include_virsorter=FALSE) {

@@

  if (include_virsorter) {
    keep_score[input_seqs$Category==1] <- keep_score[input_seqs$Category==1] + 1
    keep_score[input_seqs$Category==2] <- keep_score[input_seqs$Category==2] + 1
    keep_score[input_seqs$Category==3] <- keep_score[input_seqs$Category==3] + 0.5
    keep_score[input_seqs$Category==4] <- keep_score[input_seqs$Category==4] + 1
    keep_score[input_seqs$Category==5] <- keep_score[input_seqs$Category==5] + 1
    keep_score[input_seqs$Category==6] <- keep_score[input_seqs$Category==6] + 0.5
  }

  if (include_deepvirfinder) {
    keep_score[input_seqs$score>=0.7] <- keep_score[input_seqs$score>=0.7] + 0.5
#   keep_score[input_seqs$score>=0.9] <- keep_score[input_seqs$score>=0.9] + 0.5
  }

@@

## just kaiju
```{r}
viruses$keep_score_k <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = F,
                                            include_kaiju = T,
                                            include_deepvirfinder = F,
                                            include_virsorter = F)

@@

## deepvirfinder + checkV + kaiju
```{r}
viruses$keep_score_dvf_cv_kj <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = T,
                                            include_virsorter = F)

assess_performance(viruses$seqtype, viruses$keep_score_dvf_cv_kj)
```

@@

## just virsorter
```{r}
viruses$keep_score_vs <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = F,
                                            include_kaiju = F,
                                            include_deepvirfinder = F,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs)
```

## virsorter + checkV
```{r}
viruses$keep_score_vs_cv <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = T,
                                            include_kaiju = F,
                                            include_deepvirfinder = F,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv)
```

## virsorter + checkV + kaiju
```{r}
viruses$keep_score_vs_cv_kj <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = F,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv_kj)
```

## virsorter + checkV + kaiju + vibrant + deepvirfinder + virsorter2
```{r}
viruses$keep_score_vs_cv_kj <- getting_viral_set_1(viruses, include_vibrant = T,
                                            include_virsorter2 = T,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = T,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv_kj)
```

## virsorter + checkV + kaiju + vibrant
```{r}
viruses$keep_score_vs_cv_kj_vb <- getting_viral_set_1(viruses, include_vibrant = T,
                                            include_virsorter2 = F,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = F,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv_kj_vb)
```

## virsorter + checkV + kaiju + virsorter2 + vibrant
```{r}
viruses$keep_score_vs_cv_kj_vb_vs2 <- getting_viral_set_1(viruses, include_vibrant = T,
                                            include_virsorter2 = T,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = F,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv_kj_vb_vs2)
```

## virsorter + checkV + kaiju + virsorter2
```{r}
viruses$keep_score_vs_cv_kj_vs2_dvf <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = T,
                                            include_checkV = T,
                                            include_kaiju = T,
                                            include_deepvirfinder = T,
                                            include_virsorter = T)

assess_performance(viruses$seqtype, viruses$keep_score_vs_cv_kj_vs2_dvf)
```

@@

# Using NMDS to compare the tools
get matrix want to use for beta diversity
```{r}
virus_score_matrix <- viruses %>% select(keep_score_vb, keep_score_dvf,keep_score_vs2, keep_score_cv, keep_score_vb_cv, keep_score_dvf_cv,keep_score_vs2_cv, keep_score_cv_kj, keep_score_all)

```{r}
virus_score_matrix <- virus_score_matrix[rowSums(virus_score_matrix)>0,]
```

```{r}
library(vegan)
```

```{r}
pal <- ggthemes::tableau_color_pal(palette="Tableau 20", type="regular")

example_NMDS <- metaMDS(t(virus_score_matrix), distance = "bray")

ss <- as.data.frame(example_NMDS$points)
ss$tool <- colnames(virus_score_matrix)

p <- ggplot() + 
  geom_point(data=ss, aes(x=MDS1,y=MDS2, color=tool, fill=tool, shape=tool, size=4)) +
   theme(panel.background = element_blank(), 
        axis.line = element_line(colour = "black"),
        panel.border = element_rect(colour = "black", fill=NA, size=1),
        axis.text=element_text(size=14, colour="black"),
        axis.title=element_text(size=16),
        legend.key=element_blank(),
        legend.position="right",
        legend.background = element_blank()) +
  guides(size=FALSE) + xlab("NMDS 1") + ylab("NMDS 2") + 
  coord_equal() +
  scale_fill_manual(name="",
                     values = alpha(rev(pal(nrow(ss))), 0.5)) +
  scale_color_manual(name="",
                     values = alpha(rev(pal(nrow(ss))), 1)) +
  scale_shape_manual(name="",
                     values = c(20,21,21,22,22,23,23,24,24))

p
```

## Looking at False Positives
```{r}
viruses_false_positive <- viruses[viruses$confusion_matrix=="false positive",]
```

```{r}
table(viruses_false_positive$provirus)
```

```{r}
table(viruses_false_positive$vs2type)
```

```{r}
table(viruses_false_positive$percent_viral>50)
```

```{r}
table(viruses_false_positive$hallmark>5)
```

```{r}
table(viruses_false_positive$len.x<=10000)
```

```{r}
table(viruses_false_positive$seqtype)
```

>>>>

## Kaiju
(parsing when reading the file in "fails" for sequences that weren't classified)
```{r}
kj_c <- read_tsv("../TestingSetOutput/metagenomic_testing_set_nrandeuk.kaiju.names.out", col_names = F)
colnames(kj_c) <- c("Classified", "Contig", "NCBI_taxon", "len", "ID_best", "IDs_all", "Seq", "Name")
kj_c <- separate(kj_c, col = Contig, into = c("seqtype", "contig"), sep="--")
kj_c <- separate(kj_c, col = Name, into = c("Kaiju_Viral","Kingdom"), sep=";")
kj_c$contig <- sub("\\.", "_", kj_c$contig)
```

@@

@@

```{r}
viruses <- full_join(x=vs2_c, y=dvf_c, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=checkV, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=vb_c, by = c("contig", "seqtype"))
viruses <- full_join(x=viruses, y=kj_c, by = c("contig", "seqtype"))
```

@@

get rid of NAs for downstream processing
```{r}
viruses$Quality[is.na(viruses$Quality)] <- 0
viruses$score[is.na(viruses$score)] <- 0
viruses$bh_pvalue[is.na(viruses$bh_pvalue)] <- 0
viruses$viral[is.na(viruses$viral)] <- 0
viruses$hallmark[is.na(viruses$hallmark)] <- 0
viruses$percent_viral[is.na(viruses$percent_viral)] <- 0
viruses$percent_unknown[is.na(viruses$percent_unknown)] <- 0
viruses$Kaiju_Viral[is.na(viruses$Kaiju_Viral)] <- "unknown"
viruses$Kingdom[is.na(viruses$Kingdom)] <- "unknown"
```

@@

```{r}
getting_viral_set_1 <- function(input_seqs, include_vibrant, include_virsorter2,
                              include_deepvirfinder, include_checkV,
                              include_kaiju) {

@@

  if (include_deepvirfinder) {
    keep_score[input_seqs$score>=0.7] <- keep_score[input_seqs$score>=0.7] + 0.5
    keep_score[input_seqs$score>=0.9] <- keep_score[input_seqs$score>=0.9] + 0.5
  }

@@

## just kaiju
```{r}
viruses$keep_score_k <- getting_viral_set_1(viruses, include_vibrant = F,
                                            include_virsorter2 = F,
                                            include_checkV = F,
                                            include_kaiju = T,
                                            include_deepvirfinder = F)

@@

@@

@@
